---
title: "Coloring in Cartographic Design"
format: html
editor: visual
---

### Welcome to today's lecture!

We will talk about several different coloring styles and techniques with our familiar Oahu Median Income dataset

#### Read In Libraries

```{r}
library(tidyverse)
library(ggplot2)
library(here)
library(RColorBrewer)
library(viridisLite)
library(tidycensus)
library(mapview)
library(classInt)
library(tmap)
library(shiny)
```

#### Load in Data

```{r}
oahu_median_income <- get_acs(
  geography = "tract",
  variables = "B19013_001",  # median income
  state = "HI",
  county = "Honolulu",
  year = 2023,
  geometry = TRUE
) |> 
  dplyr::filter(GEOID != "15003981200")
```

#### Histogram of income distribution

```{r}
ggplot(data = oahu_median_income, aes(x = estimate)) +
  geom_histogram(color = "black", fill = "steelblue") +
  theme_bw()
```

#### Default Mapping

```{r}
mapview(oahu_median_income, zcol = "estimate")
```

#### Color Design

Jenks natural breaks (also called the *Jenks optimization method*) is a classification algorithm that groups continuous data into classes by minimizing the variance *within* each class and maximizing the variance *between* classes. In plain terms: it looks for natural "gaps" or clusters in the data and draws the break points there, rather than dividing the range evenly

```{r}
brks_jenks <- classIntervals(oahu_median_income$estimate, n = 6, style = "jenks")$brks

```

```{r}
brks_jenks
```

```{r}
ggplot(data = oahu_median_income, aes(x = estimate)) +
  geom_histogram(binwidth = 10000, fill = "steelblue", color = "white") +
  geom_vline(xintercept = brks_jenks, color = "firebrick", linetype = "dashed", linewidth = 0.6) +
  theme_minimal() +
  labs(x = "Median Household Income", y = "Number of Tracts",
       title = "Distribution of Median Household Income — Oahu Census Tracts",
       subtitle = "Dashed lines: Jenks natural breaks (n = 6)")
```

```{r}
mapview(oahu_median_income, zcol = "estimate", col.regions = viridis::viridis(6), at = brks_jenks)
```

We can do the same thing more simply with the `tmap` package

```{r}
tmap_mode("view")  # interactive, like mapview. Use "plot" for static.

tm_shape(oahu_median_income) +
  tm_polygons(
    "estimate",
    style = "jenks",      # <-- classification method lives right here
    n = 6,                 # <-- number of classes
    palette = "viridis",   # or "YlGnBu", "-viridis" to reverse, etc.
    title = "Median Income"
  )
```

There are many ways to segment continuous data for a color scale.

#### Classification Styles (`style =`)

| Style | What it does | Best used when |
|----|----|----|
| `pretty` | Rounds breaks to nice, readable numbers (e.g. 0, 50k, 100k) | Default choice for general audiences; readability over precision |
| `equal` | Divides the data range into equal-width bins | Data is roughly uniform/normal, no major skew or outliers |
| `quantile` | Divides data so each bin has an equal *count* of observations | You want every color used evenly; showing relative rank/percentile |
| `jenks` | Finds natural gaps in the data (minimizes within-class variance) | Skewed, real-world data (income, population); general-purpose "best default" |
| `fisher` | Similar goal to jenks (variance minimization), faster algorithm for large datasets | Same use case as jenks, but with large `n` (thousands of observations) |
| `sd` | Bins centered on the mean, in increments of standard deviation | Data represents *deviation from a mean* (e.g. anomalies, diverging data) — not raw levels |
| `kmeans` | Uses k-means clustering to find groups | Data has genuinely distinct clusters you want statistically identified |
| `hclust` | Hierarchical clustering to find groups | Similar to kmeans; useful when cluster structure may be nested/hierarchical |
| `bclust` | Bagged clustering (more robust/stable version of hclust) | Same as hclust, when you want more stability across resampling |
| `dpih` | Uses histogram bin-width optimization (plug-in bandwidth) | Statistically-driven bin width, less common for choropleths |
| `headtails` | Splits repeatedly at the mean, tailored for heavy-tailed distributions | Highly skewed data with a long tail (e.g. wealth, city sizes) |

#### Shiny App

Let's build a shiny app to compare styles

```{r}

#| label: tmap-style-explorer
#| standalone: true

ui <- fluidPage(
  titlePanel("tmap Classification Styles: Oahu Median Household Income"),
  
  sidebarLayout(
    sidebarPanel(
      selectInput(
        "style", "Classification style (style =)",
        choices = c("pretty", "equal", "quantile", "jenks", "fisher",
                    "sd", "kmeans", "hclust", "bclust", "dpih", "headtails"),
        selected = "jenks"
      ),
      
      sliderInput("n", "Number of classes (n)", min = 3, max = 9, value = 5, step = 1),
      
      selectInput(
        "palette", "Palette",
        choices = c("viridis", "magma", "plasma",
                    "YlGnBu (Brewer)" = "YlGnBu",
                    "YlOrRd (Brewer)" = "YlOrRd",
                    "BuPu (Brewer)" = "BuPu"),
        selected = "viridis"
      ),
      
      checkboxInput("reverse", "Reverse palette", value = FALSE),
      
      hr(),
      verbatimTextOutput("breaks_out")
    ),
    
    mainPanel(
      tmapOutput("map", height = "650px")
    )
  )
)

server <- function(input, output, session) {
  
  tmap_mode("view")
  
  pal <- reactive({
    p <- input$palette
    if (input$reverse) paste0("-", p) else p
  })
  
  output$map <- renderTmap({
    tm_shape(oahu_median_income) +
      tm_polygons(
        "estimate",
        style   = input$style,
        n       = input$n,
        palette = pal(),
        title   = "Median\nHousehold\nIncome",
        id      = "GEOID"
      )
  })
  
  output$breaks_out <- renderPrint({
    brks <- classInt::classIntervals(
      oahu_median_income$estimate,
      n = input$n,
      style = input$style
    )
    cat("Breaks for style =", input$style, "\n\n")
    print(round(brks$brks, 0))
  })
}

shinyApp(ui, server)
```

#### Low Income

Households under \$93000 are designated as low income. Let's map this

```{r}
oahu_median_income <- oahu_median_income |> 
  mutate(low_income = ifelse(estimate < 93000, "yes", "no"))
  
```

```{r}
ggplot(data = oahu_median_income, aes(x = estimate)) +
  geom_histogram(color = "black", fill = "steelblue") +
  theme_bw() +
  geom_vline(xintercept = 93000, color = "red")
```

```{r}
mapview(oahu_median_income, zcol = "low_income", 
        col.regions = c("forestgreen", "red")) 
```
